---
title: "Top 10 Longest and Shortest Lived Presidents"
author: Michael Bentz
date: 2022-05-16
output: html_document
---

```{r setup, include = FALSE}
library(dplyr)
library(knitr)
library(lubridate)
library(plyr)
library(tidyverse)

get_mode <- function(df, column_name) {
    df_mode <- df %>%
        dplyr::count(., .data[[column_name]]) %>%
        dplyr::arrange(dplyr::desc(.data$n)) %>%
        head(., 1) %>%
        dplyr::mutate(
            mode = dplyr::case_when(
                .data$n > 1 ~ .data[[column_name]],
                TRUE ~ NA_real_
            )
        ) %>%
        dplyr::select(., mode)

    return(df_mode)
}

get_measures_for <- function(df, column_name, label) {
    nrow <- nrow(df)

    mode <- get_mode(df, column_name)

    measures_df <- df %>% 
        dplyr::mutate(
            year_of_birth = plyr::round_any(.data$year_of_birth, 50, f = floor)
        ) %>%
        dplyr::group_by(., .data$year_of_birth) %>%
        dplyr::summarize(
            label = label,
            frequency = n(),
            mean = mean(.data[[column_name]]),
            median = median(.data[[column_name]]),
            max = max(.data[[column_name]]),
            min = min(.data[[column_name]]),
            #standard_deviation = sd(.data[[column_name]])
            standard_deviation = sqrt(sum((.data[[column_name]] - mean(.data[[column_name]]))^2 / (nrow - 1)))
        ) %>%
        cbind(., mode)

    return(measures_df)
}

get_weighted_mean <- function(df1, df2) {
    df1_nrow <- nrow(df1)
    df2_nrow <- nrow(df2)

    sum_of_means <- ((df1_nrow * df1$mean) + (df2_nrow * df2$mean))
    weighted_mean <- sum_of_means / (df1_nrow + df2_nrow)

    return(weighted_mean)
}

date_format <- "%b %d,%Y"
folder_path <- "a_problem_with_presidents"

csv_file <- list.files(path = folder_path, pattern = ".csv")
presidents <- read_csv(paste0(folder_path, "/", csv_file))
```

```{r, echo = FALSE}
cleaned_presidents <- presidents %>%
    dplyr::rename_all(., list(~ make.names(., unique = TRUE))) %>% # clean column names replacing space (' ') with .
    dplyr::rename_all(., list(~ tolower(.))) %>% # make column names lowercase
    dplyr::rename_all(., list(~ gsub("\\.", "_", .))) %>% # replace period (.) with underscore (_)
    dplyr::mutate(
        birth_date = lubridate::as_date(.data$birth_date, format = date_format), # change to date type
        death_date = lubridate::as_date(.data$death_date, format = date_format) # change to date type
    )

# exclude currently living presidents
days_lived_presidents <- cleaned_presidents %>%
    dplyr::mutate(
        # interval = lubridate::interval(.data$birth_date, replace_na(.data$death_date, lubridate::today())) # precalculate interval
        interval = lubridate::interval(.data$birth_date, .data$death_date) # precalculate interval
    ) %>%
    dplyr::mutate(
        year_of_birth = lubridate::year(.data$birth_date),
        lived_years = .data$interval %/% years(1),
        lived_months = .data$interval %/% months(1),
        lived_days = .data$interval %/% days(1)
    ) %>%
    dplyr::filter(., !is.na(.data$lived_days)) %>%
    dplyr::select(
        ., -c("interval")
    )
```

<br />

To maintain data integrity, currently living presidents have been excluded in the following report. This report showcases the top 10 longest lived presidents, and the top 10 shortest lived presidents, measured in days. The 1800s has the highest concentration of shortest lived presidents and that 1700s and 1900s have the highest concentration of longest lived presidents. These findings are demonstrated in [Figure 3](#analysis).

<br />

## Top 10

```{r, echo = FALSE}
# 10 longest lived
longest_lived_presidents <- days_lived_presidents %>%
    dplyr::arrange(dplyr::desc(lived_days)) %>%
    head(., 10)


knitr::kable(
    longest_lived_presidents, 
    col.names = gsub("[.]", " ", names(longest_lived_presidents)),
    caption = "Table 1: Top 10 presidents from longest lived to shortest lived"
)
```

<br />

```{r, echo = FALSE}
# 10 shortest lived
shortest_lived_presidents <- days_lived_presidents %>%
    dplyr::arrange(lived_days) %>%
    head(., 10)

knitr::kable(
    shortest_lived_presidents, 
    col.names = gsub("[.]", " ", names(longest_lived_presidents)),
    caption = "Table 2: Top 10 presidents from shortest lived to longest lived"
)

```


```{r, echo = FALSE, fig.cap = "Figure 1: Top 10 longest lived presidents. The 1700's and 1900's have the largest concentration of longest lived presidents."}
# plot the 10 longest lived presidents
top_10_longest_plot <- ggplot(data = longest_lived_presidents, mapping = aes(x = year_of_birth, y = lived_years)) +
    geom_point()
#top_10_longest_plot + labs(title = "Top 10 Presidents from Longest Lived to Shortest Lived", x = "Birth year", y = "Lived years") #+
```

```{r, echo = FALSE, fig.cap = "Figure 2: Top 10 shortest lived presidents. The 1800's have the largest concentration of shortest lived presidents."}


# plot the 10 shortest lived presidents
top_10_shortest_plot <- ggplot(data = shortest_lived_presidents, mapping = aes(x = year_of_birth, y = lived_years)) +
    geom_point()

#top_10_shortest_plot + labs(title = "Top 10 Presidents from Shortest Lived to Longest Lived", x = "Birth year", y = "Lived years") #+
    #theme_classic()
```

<br />

## Analysis
The birth year was rounded down to reduce noise. Rounding down to the nearest 50 year interval struck the best balance between data point volume and visual clarity.  At first glance, the left plot of Figure 3 might give the impression that shortest lived presidents are almost evenly distributed from 1750 to 1900. However, looking more closely we notice that both have a standard deviation of 0 because only a single entry exists. This is verified by the plot to the right. 


```{r, echo = FALSE}
longest_lived_measures <- get_measures_for(longest_lived_presidents, "lived_days", "longest_lived")
shortest_lived_measures <- get_measures_for(shortest_lived_presidents, "lived_days", "shortest_lived")

# weighted mean
weighted_mean <- get_weighted_mean(longest_lived_measures, shortest_lived_measures)

# apply weighted means
longest_lived_data_set <- longest_lived_measures %>%
    cbind(., weighted_mean)

shortest_lived_data_set <- shortest_lived_measures %>%
    cbind(., weighted_mean)

# combine data sets
days_lived_data_set <- longest_lived_data_set %>%
    rbind(., shortest_lived_data_set)

knitr::kable(
    days_lived_data_set, 
    col.names = gsub("[.]", " ", names(days_lived_data_set)),
    caption = "Table 3: The frequency, mean, median, max, min, standard deviation, mode, and weighted mean of the top 10 longest and shortest days lived"
)
```

<br />

```{r, echo = FALSE, figures-side, fig.show = "hold", out.width = "51%", fig.cap = "Figure 3: Top 10 longest and shortest lived presidents rounded down to the nearest 50 year interval. The concentration of longest lived days is highest during the 1701 and 1900's, whereas the concentration of shortest lived days is highest during the 1801's."}

plot <- ggplot(data = days_lived_data_set, mapping = aes(x = year_of_birth, y = mean, fill = label, color = label)) + # , y = year_of_birth)) +
    geom_bar(
        stat = "identity",
        position = position_dodge(),
        width = 30,
    ) +
    geom_point(
        show.legend = FALSE,
        position = position_dodge(30),
    ) +
    geom_errorbar(
        aes(
            ymin = mean - standard_deviation,
            ymax = mean + standard_deviation
        ),
        color = "black",
        width = 1,
        position = position_dodge(30),
    )

plot + labs(x = "Top 10", y = "Mean days")


count_plot <- ggplot(data = days_lived_data_set, mapping = aes(x = year_of_birth, y = frequency, fill = label, color = label)) + # , y = year_of_birth)) +
    geom_smooth(
        stat = "identity"
    )
count_plot + labs(x = "Top 10", y = "Frequency")
```

<br />

### Conclusion
We can clearly see that there is a correlation between presidents born in the 1800's and being listed in the top 10 shortest lived. 80% (8 out of 10) of presidents that are in the 10 top shortest lived were born during the 1800's. Whereas, 80% of presidents born in the 1700 and 1900s are listed in the top 10 longest lived.
